Weighted Model Counting with Conditional Weights for Bayesian Networks

### 2021-02-11 doc/paper2/paper.tex
Early draft of the third paper

\begin{abstract}
  Weighted model counting (WMC) has emerged as the unifying inference mechanism
  across many (probabilistic) domains. Encoding an inference problem as an
  instance of WMC typically necessitates adding extra literals and clauses. This
  is partly so because the predominant definition of WMC assigns weights to
  models based on weights on literals, and this severely restricts what
  probability distributions can be represented. We develop a measure-theoretic
  perspective on WMC and propose a way to encode conditional weights on literals
  analogously to conditional probabilities. This representation can be as
  succinct as standard WMC with weights on literals but can also expand as
  needed to represent probability distributions with less structure. To
  demonstrate the performance benefits of conditional weights over the addition
  of extra literals, we develop a new WMC encoding for Bayesian networks and
  adapt a state-of-the-art WMC algorithm \textsf{ADDMC} to the new format. Our
  experiments show that the new encoding results in significantly faster
  inference on most benchmark instances. Furthermore, the same idea can be
  adapted to other WMC algorithms and other problem domains.
\end{abstract}

\section{Introduction}

@@

In this section, we give a brief introduction to two alternative ways to think
about logical constructs such as models and formulas. Let us consider a simple
example of a propositional logic $\mathcal{L}$ with only two atoms $a$ and $b$,
and let $U = \{ a, b \}$. Then $2^U$, the power set of $U$, is the set of all
models of $\mathcal{L}$, and $2^{2^U}$ is the set of all formulas. These sets
can also be represented as Boolean algebras (e.g., using the syntax $(2^{2^U},
\land, \lor, \neg, \bot, \top)$) with a partial order $\le$ that corresponds to
set inclusion $\subseteq$---see \cref{tbl:notation_example} for examples of how
various elements can be represented in both notations. Most importantly, note
that the word \emph{atom} has completely different meanings in logic and in
Boolean algebras. An atom in $\mathcal{L}$ is an atomic formula, i.e., an
element of $U$, whereas an atom in a Boolean algebra is (in set-theoretic terms)
a singleton set. For instance, an atom in $2^{2^U}$ corresponds to a model of
$\mathcal{L}$, i.e., an element of $2^U$. Unless referring specifically to a
logic, we will use the algebraic definition of an atom and refer to logical
atoms as \emph{variables}. In the rest of the paper, for any set $U$, we will
use set-theoretic notation for $2^U$ and Boolean-algebraic notation for
$2^{2^U}$, except for (Boolean) atoms in $2^{2^U}$ that are denoted as $\{x\}$
for some model $x \in 2^U$.

\subsection{The Space of Functions on Boolean Algebras}
% TODO: it would be easier to talk about pseudo-Boolean functions instead, but
% maybe it's not necessary.

@@

Since many measures of interest may not be factorable, a well-known way to
encode them into instances of WMC is by adding more literals
\citep{DBLP:journals/ai/ChaviraD08}. We can use the measure-theoretic
perspective on WMC to show that this is always possible, however, as ensuing
sections will demonstrate, it often makes the inference task much harder in
practice.\footnote{The proofs of this and other theoretical results can be found
  in the appendix.}

@@

In this section, we describe a way to encode Bayesian networks into WMC without
restricting oneself to factorable measures and thus having to add extra
variables. We will refer to it as \texttt{cw}.

A Bayesian network is a directed acyclic graph with random variables as vertices
that defines a probability distribution over them. Let $\mathcal{V}$ denote this
set of random variables. For any random variable $X \in \mathcal{V}$, let $\im
X$ denote its set of values and $\mathrm{pa}(X)$ its set of parents. The
full probability distribution is then equal to $\prod_{X \in \mathcal{V}} \Pr(X
\mid \mathrm{pa}(X))$ \citep{DBLP:books/daglib/0023820}. For discrete Bayesian
networks (and we only consider discrete networks in this paper), each factor of
this product can be represented by a CPT. See \cref{fig:example_bn} for an
example Bayesian network that we will refer to throughout this section. For
this network, $\mathcal{V} = \{ W, F, T \}$, $\mathrm{pa}(W) = \emptyset$,
$\mathrm{pa}(F) = \mathrm{pa}(T) = \{ W \}$, $\im W = \im F = \{0, 1 \}$, and
$\im T = \{ l, m, h \}$.

@@

  On the other hand, if $X$ is not binary, we represent $X$ with $|\im X|$
  indicator variables, one for each value. We let
  \[
    \mathcal{E}(X) = \begin{cases}
      \{ \lambda_{X=1} \} & \text{if } |\im X| = 2 \\
      \{ \lambda_{X=x} \mid x \in \im X \} & \text{otherwise.}
    \end{cases}
  \]
  denote the set of indicator variables for $X$ and $\mathcal{E}^*(X) =
  \mathcal{E}(X) \cup \bigcup_{Y \in \mathrm{pa}(X)} \mathcal{E}(Y)$ denote the
  set of indicator variables for $X$ and its parents in the Bayesian network.
  Finally, let $U = \bigcup_{X \in \mathcal{V}} \mathcal{E}(X)$ denote the set
  of all indicator variables for all random variables in the Bayesian network.
\end{definition}

For example, in the Bayesian network from \cref{fig:example_bn},
$\mathcal{E}^*(T) = \{ \lambda_{T=l}, \lambda_{T=m}, \lambda_{T=h},
\lambda_{W=1} \}$.

@@

\Cref{alg:encoding} produces a function with a Boolean algebra as its domain.
This function can be represented by an ADD
\citep{DBLP:journals/fmsd/BaharFGHMPS97}. The core of \textsf{ADDMC} works by
taking an ADD $\psi\colon 2^{U} \to \mathbb{R}_{\ge 0}$ (expressed as a product
of smaller ADDs) and returning $(\exists_U\psi)(\emptyset)$
\citep{DBLP:conf/aaai/DudekPV20}. In this section, we prove that the function
$\phi$ produced by \cref{alg:encoding} can be used by \textsf{ADDMC} to
correctly compute any marginal probability of the Bayesian network that was
encoded as $\phi$.\footnote{Note that it can just as well compute \emph{any}
  probability expressed using the random variables in $\mathcal{V}$.} We begin
with \cref{lemma:cpt} which shows that any conditional weight function produces
the right answer when given a valid encoding of variable-value assignments
relevant to the CPT.

@@

First, \textsf{ADDMC} constructs the \emph{primal} (a.k.a. Gaifman) \emph{graph}
of the input CNF formula as an aid for the algorithm's heuristics. This graph
has as vertices the variables of the formula, and there is an edge between two
variables $u$ and $v$ if there is a clause in the formula that contains both $u$
and $v$. We extend this definition to functions on Boolean algebras, i.e., the
factors of $\phi$. For any pair of distinct variables $u, v \in U$, we draw an
edge between them in the primal graph if there is a function $\alpha\colon 2^X
\to \mathbb{R}_{\ge 0}$ that is a factor of $\phi$ such that $u \in X$ and $v
\in X$. For instance, a factor such as $\mathrm{CPT}_X$ will enable edges
between all distinct pairs of variables in $\mathcal{E}^*(X)$. Second, even
though the function $\phi$ produced by \cref{alg:encoding} is constructed to
have $2^U$ as its domain, sometimes the domain is effectively reduced to $2^V$
for some $V \subset U$ by the ADD manipulation algorithms that optimise the ADD
representation of a function. For a simple example, consider $\alpha: 2^{\{a\}}
\to \mathbb{R}_{\ge 0}$ defined as $\alpha(\{a\}) = \alpha(\emptyset) = 0.5$.
Then $\alpha$ can be reduced to $\alpha'\colon 2^{\emptyset} \to \mathbb{R}_{\ge
  0}$ defined as $\alpha'(\emptyset) = 0.5$. To compensate for these reductions,
for the original WMC format with a weight function $w\colon U \cup \{ \neg u
\mid u \in U \} \to \mathbb{R}_{\ge 0}$, \textsf{ADDMC} would multiply its
computed answer by $\prod_{u \in U \setminus V} w(u) + w(\neg u)$. With the new
WMC format, we instead multiply the answer by $2^{|U \setminus V|}$. Each
`excluded' variable $u \in U \setminus V$ satisfies two properties: all weights
associated with $u$ are equal to $0.5$ (otherwise the corresponding CPT would
depend on $u$, and $u$ would not be excluded), and all other CPTs are
independent of $u$ (or they may have a trivial dependence, where the probability
stays the same if $u$ is replaced with its complement). Thus, the CPT that
corresponds to $u$ still multiplies the weight of every atom in the Boolean
algebra by $0.5$, but the number of atoms under consideration is halved. To
correct for this, we multiply the final answer by two for every $u \in U
\setminus V$.

@@

%\begin{figure*}[t] % TODO: describe this in the text
%  \centering
%  \input{melt.tex}%
%  \caption{Average encoding and inference time of each encoding with their
%    original WMC algorithm (on the left) and with \textsf{ADDMC} (on the right)
%    on \emph{solved} instances. The error bars denote \SI{10}{\percent} of
%    standard deviation.}
%  \label{fig:melted}
%\end{figure*}

% TODO: keep mentioning citations along with algorithm and encoding names

In this section, we describe an experimental study comparing the six WMC
encodings for Bayesian networks when run with both \textsf{ADDMC} and the
WMC algorithms used in the original papers.\footnote{Both
  \texttt{cd05} and \texttt{cd06} cannot be run with most WMC algorithms
  including \textsf{ADDMC} because these encodings allow for additional models
  that the WMC algorithm is supposed to ignore
  \citep{DBLP:conf/ijcai/ChaviraD05,DBLP:conf/sat/ChaviraD06}.} We compare the
encodings with respect to the total time it takes to encode a Bayesian network,
compile it or run a WMC algorithm on it, and extract the (numerical) answer.
Note that while all five papers that introduce other encodings include
experimental comparisons of encoding size, that is not feasible with
\textsf{ADDMC} as even instances that are fully solved in less than
\SI{0.1}{\second} are too big to build the full ADD within reasonable time and
memory limits. The experiments were run on a computing cluster with Intel Xeon
Gold 6138 and Intel Xeon E5-2630 processors\footnote{Each instance is run on the
  same processor for all encodings.} running Scientific~Linux~7 with a
\SI{32}{\gibi\byte} memory limit and a \SI{1000}{\second} timeout on both
encoding and inference. For inference, we use \textsf{Ace} for \texttt{cd05},
\texttt{cd06}, and \texttt{d02};
\textsf{Cachet}\footnote{\url{https://cs.rochester.edu/u/kautz/Cachet/}}
\citep{DBLP:conf/sat/SangBBKP04} for \texttt{sbk05}; and \textsf{c2d}
\citep{DBLP:conf/ecai/Darwiche04} for compilation and \textsf{query-dnnf}
\footnote{\url{http://www.cril.univ-artois.fr/kc/d-DNNF-reasoner.html}} for
answer computation for \texttt{bklm16}. For encoding, we use \textsf{bn2cnf}
\footnote{\url{http://www.cril.univ-artois.fr/KC/bn2cnf.html}} for
\texttt{bklm16}, and \textsf{Ace} for all other encodings (except for
\texttt{cw}, which is implemented in Python).

\paragraph{Workarounds.} \textsf{Ace} was not used to encode evidence, as
preliminary experiments revealed that the evidence-encoding implementation
contains bugs that can lead to incorrect answers or a Java exception being
thrown on some instances of the data set (and the source code is not publicly
available). Instead, we simply list all the evidence as additional clauses in
the encoding. Furthermore, to ensure that \texttt{bklm16} (whether run with
\textsf{ADDMC} or \textsf{c2d}) returns correct answers on most instances, we
had to disable one of the improvements that \texttt{bklm16} brings over
\texttt{cd06}, namely, the construction of a scaling factor that `absorbs' one
probability from each CDT \citep{DBLP:conf/ecai/BartKLM16}. For realistic
benchmark instances, this scaling factor can easily be below $10^{-30}$, and
thus would require arbitrary-precision floating-point arithmetic to be usable.
Even a toy Bayesian network with seven binary independent variables with
probabilities $0.1$ and $0.9$ is enough for \textsf{bn2cnf} to output precisely
zero as the scaling factor. We note that this issue likely remained unnoticed
because \citet{DBLP:conf/ecai/BartKLM16} did not attempt to compute numerical
answers in their experiments.

@@

\paragraph{Instances.} For experimental data, we use the Bayesian networks
available with \textsf{Ace} and \textsf{Cachet}, most of which happen to be
binary. We classify them into the following seven categories:
\begin{itemize*}
\item DQMR and
\item Grid networks as described by \citet{DBLP:conf/aaai/SangBK05},
\item Mastermind, and
\item Random Blocks from the work of \citet{DBLP:journals/ijar/ChaviraDJ06},
\item remaining binary Bayesian networks that include Plan Recognition
  \citep{DBLP:conf/aaai/SangBK05}, Friends and Smokers, Students and Professors
  \citep{DBLP:journals/ijar/ChaviraDJ06}, and \texttt{tcc4f}, and
\item non-binary classic Bayesian networks (\texttt{alarm}, \texttt{diabetes},
  \texttt{hailfinder}, \texttt{mildew}, \texttt{munin1}--\texttt{4},
  \texttt{pathfinder}, \texttt{pigs}, \texttt{water}).
\end{itemize*}
We run \textsf{ADDMC} with each of the five encodings once on each Bayesian
network.

@@

% TODO: need to update
\paragraph{Results.} \Cref{fig:cumulative} shows that \texttt{cd05} and
\texttt{cd06} (when run with \textsf{Ace}) are in the lead, while \textsf{ADDMC}
significantly underperforms when combined with any of the previous encodings.
Our encoding \texttt{cw} significantly improves the performance of
\textsf{ADDMC}, making $\textsf{ADDMC}+\texttt{cw}$ comparable to
$\textsf{Ace}+\texttt{d02}$, $\textsf{c2d}+\texttt{bklm16}$, and
$\textsf{Cachet}+\texttt{sbk05}$. Furthermore,
\cref{tbl:tallies} shows that, while $\textsf{Ace}+\texttt{cd06}$ managed to
solve the most instances, $\textsf{ADDMC}+\texttt{cw}$ was the best-performing
algorithm-encoding combination on the largest number of instances. The scatter
plot on the left hand side of \cref{fig:scatter} add to this by showing that
\texttt{cw} is particularly promising on Grid networks and tackles all DQMR
instances in less than a second. The scatter plot on the right hand side of
\cref{fig:scatter} shows that \texttt{cw} is better than \texttt{sbk05} (i.e.,
the second best performing encoding for \textsf{ADDMC}) on the majority of
instances. Seeing how, e.g., DQMR instances are trivial for
$\textsf{ADDMC}+\texttt{cw}$ but hard for $\textsf{Ace}+\texttt{cd06}$, and
vice versa for Mastermind instances, we conclude that the best-performing
algorithm-encoding combination depends significantly on (as-of-yet unknown)
properties of the Bayesian networks.

@@

We attempt to explain what makes \textsf{ADDMC} run hundreds of times faster
with \texttt{cw} than with any other encoding by considering asymptotic upper
bounds on the numbers of variables and ADDs based on the size and structure of
the Bayesian network. Let $n = |\mathcal{V}|$ be the number of vertices in the
Bayesian network, $d = \max_{X \in \mathcal{V}} |\mathrm{pa}(X)|$ the maximum
in-degree (i.e., the number of parents), and $v = \max_{X \in \mathcal{V}} |\im
X|$ the maximum number of values per variable. \Cref{tbl:asymptotes} shows how
\texttt{cw} has fewer variables and fewer ADDs than any other encoding. We
conjecture that it is primarily the reduced number of variables that makes the
\textsf{ADDMC} variable ordering heuristics much more effective. Note that these
are upper bounds and most encodings (including \texttt{cw}) can be smaller in
certain situations (e.g., with binary random variables or when a CPT has
repeating probabilities). We equate clauses and ADDs (more specifically, factors
of the function $\phi$ from \cref{alg:encoding}) here because \textsf{ADDMC}
interprets each clause of any WMC encoding as a multiplicative factor of the ADD
that represents the entire WMC instance \citep{DBLP:conf/aaai/DudekPV20}. For
literal-weight encodings, each weight is also a factor, but that does not affect
the asymptotic bounds in the table.

\section{Discussion and Conclusion}
% TODO: mention that DPMC is the next step
% TODO: which algorithm comes out on top depends on the data set
% TODO: emphasise that BN encodings have never been compared w.r.t. time before

WMC was originally motivated by an appeal to the success of \SAT{} solvers in
efficiently tackling an $\NP{}$-complete problem
\citep{DBLP:conf/aaai/SangBK05}. \textsf{ADDMC} does not rely on \SAT{}-based
algorithmic techniques, and our proposed format diverges even more from the
DIMACS CNF format for Boolean formulas. To what extent are \SAT{}-based methods
still applicable? The answer depends significantly on the problem domain. For
Bayesian networks, the rules describing that each random variable can only be
associated with exactly one value were still encoded as clauses. As has been
noted previously \citep{DBLP:conf/sat/ChaviraD06}, rows in CPTs with
probabilities equal to zero or one can be represented as clauses as well.
Therefore, our work can be seen as proposing a middle ground between $\#\SAT{}$
and probabilistic inference.

While we chose to use \textsf{ADDMC} \citep{DBLP:conf/aaai/DudekPV20} as the WMC
algorithm and Bayesian networks as a canonical example of a probabilistic
inference task, these are only examples meant to illustrate the broader idea
that choosing a more expressive representation of weights can outperform
increasing the size of the problem to keep the weights simple. Indeed, in this
work, we have provided a new theoretical perspective on the expressive power of
WMC and illustrated the empirical benefits of that perspective. The same idea
can be adapted to other inference problem domains such as probabilistic programs
\citep{DBLP:journals/tplp/FierensBRSGTJR15,DBLP:journals/corr/abs-2005-09089} as
well as to search-based solvers such as \textsf{Cachet}
\citep{DBLP:conf/sat/SangBBKP04}, although the extension may be less
straightforward for other WMC techniques.

@@

\bibliography{paper}

>>>>

\begin{abstract}
  Weighted model counting (WMC) has emerged as the unifying inference mechanism
  across many (probabilistic) domains. Encoding an inference problem as an
  instance of WMC typically necessitates adding extra literals and clauses. This
  is partly so because the predominant definition of WMC assigns weights to
  models based on weights on literals, and this severely restricts what
  probability distributions can be represented. We develop a measure-theoretic
  perspective on WMC and propose a way to encode conditional weights on literals
  analogously to conditional probabilities. This representation can be as
  succinct as standard WMC with weights on literals but can also expand as
  needed to represent probability distributions with less structure. To
  demonstrate the performance benefits of conditional weights over the addition
  of extra literals, we develop a new WMC encoding for Bayesian networks and
  adapt a state-of-the-art WMC algorithm \textsf{ADDMC} to the new format. Our
  experiments show that the new encoding significantly improves the performance
  of the algorithm on most benchmark instances. Furthermore, the same idea can
  be adapted to other WMC algorithms and other problem domains.
\end{abstract}

\section{Introduction}
% TODO: keep mentioning citations along with algorithm and encoding names

@@

In this section, we give a brief introduction to two alternative ways to think
about logical constructs such as models and formulas. Let us consider a simple
example of a propositional logic $\mathcal{L}$ with only two atoms $a$ and $b$,
and let $U = \{ a, b \}$. Then $2^U$, the power set of $U$, is the set of all
models of $\mathcal{L}$, and $2^{2^U}$ is the set of all formulas. These sets
can also be represented as Boolean algebras (e.g., using the syntax $(2^{2^U},
\land, \lor, \neg, \bot, \top)$) with a partial order $\le$ that corresponds to
set inclusion $\subseteq$---see \cref{tbl:notation_example} for examples of how
various elements can be represented in both notations. Most importantly, note
that the word \emph{atom} has completely different meanings in logic and Boolean
algebras. An atom in $\mathcal{L}$ is an atomic formula, i.e., an element of
$U$, whereas an atom in a Boolean algebra is (in set-theoretic terms) a
singleton set. For instance, an atom in $2^{2^U}$ corresponds to a model of
$\mathcal{L}$, i.e., an element of $2^U$. Unless referring specifically to a
logic, we will use the algebraic definition of an atom and refer to logical
atoms as \emph{variables}. In the rest of the paper, for any set $U$, we will
use set-theoretic notation for $2^U$ and Boolean-algebraic notation for
$2^{2^U}$, except for (Boolean) atoms in $2^{2^U}$ that are denoted as $\{x\}$
for some model $x \in 2^U$.

\subsection{Functions on Boolean Algebras}
% TODO: maybe: it would be easier to talk about pseudo-Boolean functions
% instead, but maybe it's not necessary.

@@

Since many measures of interest may not be factorable, a well-known way to
encode them into instances of WMC is by adding more literals
\citep{DBLP:journals/ai/ChaviraD08}. We can use the measure-theoretic
perspective on WMC to show that this is always possible, however, as ensuing
sections will demonstrate, it can make the inference task much harder in
practice.\footnote{The proofs of this and other theoretical results can be found
  in the appendix.}

@@

In this section, we describe a way to encode Bayesian networks into WMC without
restricting oneself to factorable measures and thus having to add extra
variables. We will refer to it as \texttt{cw}. A Bayesian network is a directed
acyclic graph with random variables as vertices that defines a probability
distribution over them. Let $\mathcal{V}$ denote this set of random variables.
For any random variable $X \in \mathcal{V}$, let $\im X$ denote its set of
values and $\mathrm{pa}(X)$ its set of parents. The full probability
distribution is then equal to $\prod_{X \in \mathcal{V}} \Pr(X \mid
\mathrm{pa}(X))$. For discrete Bayesian networks (and we only consider discrete
networks in this paper), each factor of this product can be represented by a
CPT. See \cref{fig:example_bn} for an example Bayesian network that we will
refer to throughout this section. For this network, $\mathcal{V} = \{ W, F, T
\}$, $\mathrm{pa}(W) = \emptyset$, $\mathrm{pa}(F) = \mathrm{pa}(T) = \{ W \}$,
$\im W = \im F = \{0, 1 \}$, and $\im T = \{ l, m, h \}$.

@@

  On the other hand, if $X$ is not binary, we represent $X$ with $|\im X|$
  indicator variables, one for each value. We let
  \[
    \mathcal{E}(X) = \begin{cases}
      \{ \lambda_{X=1} \} & \text{if } |\im X| = 2 \\
      \{ \lambda_{X=x} \mid x \in \im X \} & \text{otherwise.}
    \end{cases}
  \]
  denote the set of indicator variables for $X$ and $\mathcal{E}^*(X) =
  \mathcal{E}(X) \cup \bigcup_{Y \in \mathrm{pa}(X)} \mathcal{E}(Y)$ denote the
  set of indicator variables for $X$ and its parents in the Bayesian network.
  Finally, let $U = \bigcup_{X \in \mathcal{V}} \mathcal{E}(X)$ denote the set
  of all indicator variables for all random variables in the Bayesian network.
  For example, in the Bayesian network from \cref{fig:example_bn},
  $\mathcal{E}^*(T) = \{ \lambda_{T=l}, \lambda_{T=m}, \lambda_{T=h},
  \lambda_{W=1} \}$.
\end{definition}

@@

\Cref{alg:encoding} produces a function with a Boolean algebra as its domain.
This function can be represented by an ADD
\citep{DBLP:journals/fmsd/BaharFGHMPS97}. \textsf{ADDMC} takes an ADD
$\psi\colon 2^{U} \to \mathbb{R}_{\ge 0}$ (expressed as a product of smaller
ADDs) and returns $(\exists_U\psi)(\emptyset)$ \citep{DBLP:conf/aaai/DudekPV20}.
In this section, we prove that the function $\phi$ produced by
\cref{alg:encoding} can be used by \textsf{ADDMC} to correctly compute any
marginal probability of the Bayesian network that was encoded as
$\phi$.\footnote{Note that it can just as well compute \emph{any} probability
  expressed using the random variables in $\mathcal{V}$.} We begin with
\cref{lemma:cpt} which shows that any conditional weight function produces the
right answer when given a valid encoding of variable-value assignments relevant
to the CPT.

@@

First, \textsf{ADDMC} constructs the \emph{primal} (a.k.a. Gaifman) \emph{graph}
of the input CNF formula as an aid for the algorithm's heuristics. This graph
has as vertices the variables of the formula, and there is an edge between two
variables $u$ and $v$ if there is a clause in the formula that contains both $u$
and $v$. We extend this definition to functions on Boolean algebras, i.e., the
factors of $\phi$. For any pair of distinct variables $u, v \in U$, we draw an
edge between them in the primal graph if there is a function $\alpha\colon 2^X
\to \mathbb{R}_{\ge 0}$ that is a factor of $\phi$ such that $u, v \in X$. For
instance, a factor such as $\mathrm{CPT}_X$ will enable edges between all
distinct pairs of variables in $\mathcal{E}^*(X)$. Second, even though the
function $\phi$ produced by \cref{alg:encoding} is constructed to have $2^U$ as
its domain, sometimes the domain is effectively reduced to $2^V$ for some $V
\subset U$ by the ADD manipulation algorithms that optimise the ADD
representation of a function. For a simple example, consider $\alpha: 2^{\{a\}}
\to \mathbb{R}_{\ge 0}$ defined as $\alpha(\{a\}) = \alpha(\emptyset) = 0.5$.
Then $\alpha$ can be reduced to $\alpha'\colon 2^{\emptyset} \to \mathbb{R}_{\ge
  0}$ defined as $\alpha'(\emptyset) = 0.5$. To compensate for these reductions,
for the original WMC format with a weight function $w\colon U \cup \{ \neg u
\mid u \in U \} \to \mathbb{R}_{\ge 0}$, \textsf{ADDMC} would multiply its
computed answer by $\prod_{u \in U \setminus V} w(u) + w(\neg u)$. With the new
WMC format, we instead multiply the answer by $2^{|U \setminus V|}$. Each
`excluded' variable $u \in U \setminus V$ satisfies two properties: all weights
associated with $u$ are equal to $0.5$ (otherwise the corresponding CPT would
depend on $u$, and $u$ would not be excluded), and all other CPTs are
independent of $u$ (or they may have a trivial dependence, where the probability
stays the same if $u$ is replaced with its complement). Thus, the CPT that
corresponds to $u$ still multiplies the weight of every atom in the Boolean
algebra by $0.5$, but the number of atoms under consideration is halved. To
correct for this, we multiply the final answer by two for every $u \in U
\setminus V$.

@@

%\begin{figure*}[t] % TODO: maybe not: describe this in the text
%  \centering
%  \input{melt.tex}%
%  \caption{Average encoding and inference time of each encoding with their
%    original WMC algorithm (on the left) and with \textsf{ADDMC} (on the right)
%    on \emph{solved} instances. The error bars denote \SI{10}{\percent} of
%    standard deviation.}
%  \label{fig:melted}
%\end{figure*}

We compare the six WMC encodings for Bayesian networks when run with both
\textsf{ADDMC} and the WMC algorithms used in the original papers.\footnote{Both
  \texttt{cd05} and \texttt{cd06} cannot be run with most WMC algorithms
  including \textsf{ADDMC} because these encodings allow for additional models
  that the WMC algorithm is supposed to ignore
  \citep{DBLP:conf/ijcai/ChaviraD05,DBLP:conf/sat/ChaviraD06}.} We compare the
encodings with respect to the total time it takes to encode a Bayesian network,
compile it or run a WMC algorithm on it, and extract the (numerical) answer.
Note that while all five papers that introduce other encodings include
experimental comparisons of encoding size, that is not feasible with
\textsf{ADDMC} as even instances that are fully solved in less than
\SI{0.1}{\second} are too big to build the full ADD within reasonable time and
memory limits. The experiments were run on a computing cluster with Intel Xeon
Gold 6138 and Intel Xeon E5-2630 processors\footnote{Each instance is run on the
  same processor for all encodings.} running Scientific~Linux~7 with a
\SI{32}{\gibi\byte} memory limit and a \SI{1000}{\second} timeout on both
encoding and inference. For inference, we use \textsf{Ace} for \texttt{cd05},
\texttt{cd06}, and \texttt{d02};
\textsf{Cachet}\footnote{\url{https://cs.rochester.edu/u/kautz/Cachet/}}
\citep{DBLP:conf/sat/SangBBKP04} for \texttt{sbk05}; and \textsf{c2d}
\citep{DBLP:conf/ecai/Darwiche04} for compilation and \textsf{query-dnnf}
\footnote{\url{http://www.cril.univ-artois.fr/kc/d-DNNF-reasoner.html}} for
answer computation for \texttt{bklm16}. For encoding, we use \textsf{bn2cnf}
\footnote{\url{http://www.cril.univ-artois.fr/KC/bn2cnf.html}} for
\texttt{bklm16}, and \textsf{Ace} for all other encodings (except for
\texttt{cw}, which is implemented in Python).

\textsf{Ace} was not used to encode evidence, as
preliminary experiments revealed that the evidence-encoding implementation
contains bugs that can lead to incorrect answers or a Java exception being
thrown on some instances of the data set (and the source code is not publicly
available). Instead, we simply list all the evidence as additional clauses in
the encoding. Furthermore, to ensure that \texttt{bklm16} (whether run with
\textsf{ADDMC} or \textsf{c2d}) returns correct answers on most instances, we
had to disable one of the improvements that \texttt{bklm16} brings over
\texttt{cd06}, namely, the construction of a scaling factor that `absorbs' one
probability from each CDT \citep{DBLP:conf/ecai/BartKLM16}. For realistic
benchmark instances, this scaling factor can easily be below $10^{-30}$, and
thus would require arbitrary-precision floating-point arithmetic to be usable.
Even a toy Bayesian network with seven binary independent variables with
probabilities $0.1$ and $0.9$ is enough for \textsf{bn2cnf} to output precisely
zero as the scaling factor. We note that this issue likely remained unnoticed
because \citet{DBLP:conf/ecai/BartKLM16} did not attempt to compute numerical
answers in their experiments.

@@

For experimental data, we use the Bayesian networks
available with \textsf{Ace} and \textsf{Cachet}, most of which happen to be
binary. We classify them into the following seven categories:
\begin{itemize*}
\item DQMR and
\item Grid networks as described by \citet{DBLP:conf/aaai/SangBK05},
\item Mastermind, and
\item Random Blocks from the work of \citet{DBLP:journals/ijar/ChaviraDJ06},
\item remaining binary Bayesian networks that include Plan Recognition
  \citep{DBLP:conf/aaai/SangBK05}, Friends and Smokers, Students and Professors
  \citep{DBLP:journals/ijar/ChaviraDJ06}, and \texttt{tcc4f}, and
\item non-binary classic Bayesian networks (\texttt{alarm}, \texttt{diabetes},
  \texttt{hailfinder}, \texttt{mildew}, \texttt{munin1}--\texttt{4},
  \texttt{pathfinder}, \texttt{pigs}, \texttt{water}).
\end{itemize*}
We run \textsf{ADDMC} with each of the five encodings once on each Bayesian
network.

@@

\Cref{fig:cumulative} shows that \texttt{cd05} and
\texttt{cd06} (when run with \textsf{Ace}) are in the lead, while \textsf{ADDMC}
significantly underperforms when combined with any of the previous encodings.
Our encoding \texttt{cw} significantly improves the performance of
\textsf{ADDMC}, making $\textsf{ADDMC}+\texttt{cw}$ comparable to
$\textsf{Ace}+\texttt{d02}$, $\textsf{c2d}+\texttt{bklm16}$, and
$\textsf{Cachet}+\texttt{sbk05}$. Furthermore,
\cref{tbl:tallies} shows that, while $\textsf{Ace}+\texttt{cd06}$ managed to
solve the most instances, $\textsf{ADDMC}+\texttt{cw}$ was the best-performing
algorithm-encoding combination on the largest number of instances. The scatter
plot on the left-hand side of \cref{fig:scatter} add to this by showing that
\texttt{cw} is particularly promising on Grid networks and tackles all DQMR
instances in less than a second. The scatter plot on the right-hand side of
\cref{fig:scatter} shows that \texttt{cw} is better than \texttt{sbk05} (i.e.,
the second-best encoding for \textsf{ADDMC}) on the majority of
instances. Seeing how, e.g., DQMR instances are trivial for
$\textsf{ADDMC}+\texttt{cw}$ but hard for $\textsf{Ace}+\texttt{cd06}$, and
vice versa for Mastermind instances, we conclude that the best-performing
algorithm-encoding combination depends significantly on (as-of-yet unknown)
properties of the Bayesian networks.

@@

We attempt to explain what makes \textsf{ADDMC} run hundreds of times faster
with \texttt{cw} than with any other encoding by considering asymptotic upper
bounds on the numbers of variables and ADDs based on the size and structure of
the Bayesian network. Let $n = |\mathcal{V}|$ be the number of vertices in the
Bayesian network, $d = \max_{X \in \mathcal{V}} |\mathrm{pa}(X)|$ the maximum
in-degree (i.e., the number of parents), and $v = \max_{X \in \mathcal{V}} |\im
X|$ the maximum number of values per variable. \Cref{tbl:asymptotes} shows how
\texttt{cw} has fewer variables and fewer ADDs than any other encoding. We
conjecture that it is primarily the reduced number of variables that makes the
\textsf{ADDMC} variable ordering heuristics much more effective. Note that these
are upper bounds and most encodings (including \texttt{cw}) can be smaller in
certain situations (e.g., with binary random variables or when a CPT has
repeating probabilities). We equate clauses and ADDs (more specifically, factors
of the function $\phi$ from \cref{alg:encoding}) here because \textsf{ADDMC}
interprets each clause of any WMC encoding as a multiplicative factor of the ADD
that represents the entire WMC instance \citep{DBLP:conf/aaai/DudekPV20}. For
literal-weight encodings, each weight is also a factor, but that does not affect
our asymptotic bounds.

\section{Conclusions and Future Work}
% TODO: maybe: emphasise that BN encodings have never been compared w.r.t. time
% before

WMC was originally motivated by an appeal to the success of \SAT{} solvers in
efficiently tackling an $\NP{}$-complete problem
\citep{DBLP:conf/aaai/SangBK05}. \textsf{ADDMC} does not rely on \SAT{}-based
algorithmic techniques \citep{DBLP:conf/aaai/DudekPV20}, and our proposed format
diverges even more from the DIMACS CNF format for Boolean formulas. To what
extent are \SAT{}-based methods still applicable? The answer depends
significantly on the problem domain. For Bayesian networks, the rules describing
that each random variable can only be associated with exactly one value were
still encoded as clauses. As has been noted previously
\citep{DBLP:conf/sat/ChaviraD06}, rows in CPTs with probabilities equal to zero
or one can be represented as clauses as well. Therefore, our work can be seen as
proposing a middle ground between $\#\SAT{}$ and probabilistic inference.

While we chose \textsf{ADDMC} as the WMC algorithm and Bayesian networks as a
canonical example of a probabilistic inference task, these are only examples
meant to illustrate the broader idea that choosing a more expressive
representation of weights can outperform increasing the size of the problem to
keep the weights simple. Indeed, in this work, we have provided a new
theoretical perspective on the expressive power of WMC and illustrated the
empirical benefits of that perspective. The same idea can be adapted to other
inference problem domains such as probabilistic programs
\citep{DBLP:journals/tplp/FierensBRSGTJR15,DBLP:journals/corr/abs-2005-09089} as
well as to search-based solvers such as \textsf{Cachet}
\citep{DBLP:conf/sat/SangBBKP04} and \textsf{DPMC}
---an extension to \textsf{ADDMC} that adds support for computations based on
tensors (rather than ADDs) and planning based on tree decompositions
\cite{DBLP:conf/cp/DudekPV20}. Another important direction for future work is to
develop a better understanding of what properties of Bayesian networks make an
instance easy for some algorithm-encoding combinations more than others.

@@

\clearpage
\bibliography{paper}
